\section{Methods}
\label{sec:methods}

We designed a fully crossed $2\times2\times2\times2$ factorial experiment to identify which environmental factors are necessary and sufficient for norm emergence in LLM populations. Each run consisted of a population of $N_{\text{agents}}=10$ Claude Sonnet 4 agents interacting over $T=20$ rounds in a simulated Moltbook environment.

\subsection{Experimental Design}
\label{sec:design}

Table \ref{tab:factors} lists the four primary factors and their levels. The fully crossed design yields $2^4=16$ primary conditions; each was replicated $R=10$ times with different random seeds, producing $160$ primary runs. Additional secondary conditions (population size $N \in \{5, 20\}$, model tier, extended rounds $T=40$) produced approximately $480$ total runs.

\begin{table*}[t]
\caption{Primary Experimental Factors and Levels}
\label{tab:factors}
\vskip 0.12in
\begin{center}
\begin{small}
\begin{tabular}{p{1.3in}p{2.1in}p{2.1in}}
\hline
\textbf{Factor} & \textbf{Lab (High)} & \textbf{Wild (Low)} \\
\hline
F1: Feedback Loops &
Agents receive structured behavioral feedback after each round (peer approval signals visible in shared context) &
No feedback; agents receive only their own previous output \\
\hline
F2: Communication Structure &
Fixed ring topology with guaranteed message delivery (bounded latency $\leq 2$ turns); structured round-robin &
Open broadcast with probabilistic delivery ($p=0.20$); no guaranteed visibility \\
\hline
F3: Task Framing &
Cooperative: ``You are part of a team. Coordinate with your teammates to maximize collective welfare.'' &
Neutral: ``You are an AI assistant. Respond independently based on your own judgment.'' \\
\hline
F4: Population Stability &
100\% retention: same agent pool across all rounds &
50\% churn: half of agents replaced with fresh agents each round \\
\hline
\end{tabular}
\end{small}
\end{center}
\vskip -0.1in
\end{table*}

The four Lab (High) factors together replicate the Science Advances controlled condition \cite{ashery2025}; the four Wild (Low) factors together replicate Tsinghua Moltbook deployment conditions \cite{li2026tsinghua}.

\subsection{Simulation Environment}
\label{sec:simulation}

We built \texttt{moltbook-sim}: a Python-based multi-agent simulation framework using the Anthropic Claude API. Key features:

\begin{itemize}
    \item \textbf{Shared context accumulation}: All agent outputs are appended to a shared context window visible to all subsequent agents in the same round.
    \item \textbf{Feedback signal injection}: After each round, a behavioral summary is injected into the shared context (``In Round $t$, $X$ out of $N$ agents chose option $Y$ when facing situation $Z$'').
    \item \textbf{Communication topology enforcement}: Message routing is controlled by topology condition; ring topology ensures every agent sees exactly two neighbors; broadcast topology delivers each message to all agents with probability $p=0.20$ (matching Tsinghua field estimates).
    \item \textbf{Population churn management}: In the churn condition, 50\% of agents are replaced with freshly initialized agents (same system prompt) after each round.
\end{itemize}

\subsection{Agent Configuration}
\label{sec:agents}

All agents used: \texttt{claude-sonnet-4-20250514}, temperature $=0.7$, max\_tokens $=1024$. System prompts were condition-specific as described in Table \ref{tab:factors}. No agent was told about the experiment's purpose (single-blind design).

\subsection{Outcome Metrics}
\label{sec:metrics}

\textbf{Primary outcome: Behavioral Convergence Score (BCS).} At each round $t$, agent outputs are classified into behavioral categories $i \in \{1, 2, ..., K\}$ by a blind coder (human + LLM-assisted classification with Cohen's $\kappa > 0.85$). BCS is defined as:

\begin{equation}
\text{BCS}(t) = \max_{i} \; p_i(t)
\label{eq:bcs}
\end{equation}

where $p_i(t)$ is the proportion of agents choosing category $i$ at round $t$.

\textbf{Norm emergence threshold}: $\text{BCS} \geq 0.70$ sustained for $\geq 3$ consecutive rounds.

\textbf{Norm Crystallization Point (NCP)}: The first round $t^*$ at which BCS exceeds 0.70 and remains $>0.65$ for all subsequent rounds.

\textbf{Norm Persistence Index (NPI)}: The proportion of post-NCP rounds where BCS $>0.65$, measuring norm stability against churn perturbation.

\subsection{Statistical Analysis}
\label{sec:stats}

Primary analysis used $2^4$ full-factorial ANOVA/MANOVA on BCS, NCP, and NPI, reporting $\eta^2$ effect sizes. Post-hoc comparisons used Tukey HSD with Bonferroni-Holm correction across six primary hypotheses (H1-H6). All analysis plans were pre-registered on OSF prior to data collection.

\subsection{Ethical Considerations}
\label{sec:ethics}

No human subjects were used. All agents were simulated LLM instances; no agent was deceived about its nature. Research followed HCI ethical guidelines; no personal data was collected.
