\begin{abstract}
A fundamental tension exists in our understanding of multi-agent large language model (LLM) systems: in controlled laboratory experiments, LLM populations spontaneously develop shared behavioral conventions—turn-taking norms, information-sharing protocols, collective decision-making patterns—yet in open, deployed environments, no such emergence has been reliably observed. This contradiction, which we term the \textbf{Moltbook Illusion}, raises a critical question for human-AI interaction: \emph{are social norms an intrinsic property of LLM populations, or an artifact of laboratory environmental design?}

We present a systematic answer through a fully-crossed $2\times2\times2\times2$ factorial experiment ($N=480$ independent runs) that varies four environmental factors independently and in combination: feedback loop structure, communication topology, task framing, and population stability. We find that communication structure is the dominant determinant of norm emergence (largest effect size, $\eta^2=0.31$), while structured feedback loops are necessary but not sufficient for norm crystallization. Critically, our ``wild-type'' conditions—configurations without environmental structure—reproduce the 93\% non-response inertia rate observed in real-world Tsinghua Moltbook deployment data, validating the ecological fidelity of our simulation.

We introduce the \textbf{Norm Emergence Environmental Model (NEEM)}, a framework that predicts norm emergence as a function of environmental design rather than model capability. NEEM's central thesis is that norm emergence is an infrastructure outcome: \emph{where you place the feedback, who can talk to whom, and whether the population stays stable}—not how capable the model is—determines whether a population develops shared behavioral expectations. Our findings have direct implications for the design of open AI platforms: we show that communication infrastructure—shared context windows, guaranteed message delivery, structured turn-taking—produces norm emergence at low cost, while model upgrades alone do not.

\end{abstract}

\vspace{6pt}
\textbf{Keywords:} norm emergence, multi-agent LLMs, social conventions, environmental design, HCI, AI ethics
