\section{Limitations}
\label{sec:limitations}

We acknowledge four primary limitations, each with specific mitigations and directions for future work.

\subsection{Ecological Validity}
Our ``wild'' condition is an approximation of—not a direct replication of—Tsinghua Moltbook conditions. Key parameters (communication probability $p=0.20$, churn rate 50\%) are estimates derived from published deployment descriptions \citep{li2026tsinghua} rather than raw field data. We mitigate this through quantitative baseline matching: our wild condition reproduces the 93\% IIR (91.3\% observed, difference 1.7pp, 95\% CI [-3.8, 7.2]), confirming that the simulation captures the essential dynamics. We do not claim full ecological equivalence; this is a first approximation that should be validated against direct field data in future work.

\subsection{Model Dependency and Generalizability}
All experiments use Claude Sonnet 4 (claude-sonnet-4-20250514). Results may not generalize to other LLMs with different instruction-following profiles or social reasoning capabilities. Critically, \citeA{piedrahita2025free} demonstrate that reasoning-capable models can exhibit \emph{degraded} cooperative behavior, which may interact with our framing (F3) and feedback (F1) effects in ways our study cannot address. We recommend cross-model replication as essential future work before drawing strong conclusions about optimal environmental design across LLM families.

\subsection{Norm Operationalization: Convergence $\neq$ Norms}
BCS measures behavioral convergence but cannot, by itself, distinguish genuine social norms from coincidental alignment (e.g., agents arriving at similar responses independently because the task has a single correct answer). We mitigate this through three converging measures: NPI (persistence against churn), CSNT (cross-session transfer), and qualitative behavioral analysis of agent outputs. A critic might still argue that BCS-based evidence is necessary but not sufficient for claiming genuine norm emergence. We are transparent about this limitation and welcome more refined operationalizations in future work.

\subsection{Absence of Human-in-the-Loop}
This is an agent-only experiment. Human-AI mixed populations introduce dynamics our study cannot capture: human social cognition, cultural priors, strategic behavior, and real social consequences. The HCI community has long recognized that human-computer interaction differs fundamentally from computer-computer interaction. Our findings apply most directly to agent-agent systems (e.g., autonomous coding teams, research pipelines); their applicability to human-agent and human-human settings via AI mediation requires direct empirical validation.
