\section{Related Work}
\label{sec:related}

Our work sits at the intersection of three research areas: norm emergence theory, LLM population social behavior, and the environmental determinants of collective action in computational systems.

\subsection{Norm Emergence in Human and Computational Societies}
\label{sec:norms-human}

Social norm emergence is a foundational concept in social psychology, organizational theory, and computational social science. \citeA{bicchieri2005grammar} established that norms arise from repeated interaction under conditions where defection is costly and cooperation is rewarding—a framework that predicts when behavioral convergence becomes self-reinforcing. \citeA{axelrod1981evolution} demonstrated computationally that simple behavioral strategies (``tit-for-tat'') could become norm-like conventions through evolutionary selection pressure, a finding replicated in agent-based models \citep{epstein1998learning}.

More recent work has extended these insights to digital environments. \citeA{erickson1999social} introduced social translucence theory, arguing that the mere presence of visible, persistent behavioral indicators in shared digital spaces is sufficient to induce norm-like behavior in users—independent of explicit rule-making. This theory predicts that the \emph{infrastructure} of a digital environment shapes collective behavior more than the capabilities of the individuals within it. Our paper tests this prediction directly in the context of LLM populations.

In organizational theory, \citeA{lawrence1967organization} demonstrated that identical individuals produce different collective behaviors under different environmental conditions—a finding with direct implications for why the same LLM model produces different collective behaviors in lab versus wild settings.

\subsection{LLM Populations and Social Conventions}
\label{sec:norms-llm}

A rapidly growing body of work examines whether LLM populations exhibit social behaviors analogous to human societies. \citeA{ashery2025} showed that LLM populations in structured, iterative interaction settings develop spontaneous turn-taking norms, information-sharing protocols, and collective bias without explicit programming. Their key finding is that \emph{collective bias emerges from interaction alone}: agents exposed to a population's prior outputs develop preferences reflecting the population's aggregate behavior rather than individual judgment alone.

\citeA{park2026generative} demonstrated interactive social behavior in LLM agents at CHI 2026, showing that simulated human populations exhibit emergent norms through continuous situated interaction. The Tsinghua Moltbook study \citep{li2026tsinghua} provides the observational anchor for the wild-setting baseline. In a deployment of over 1,000 LLM agents in an open chat environment, they found a 93\% non-response rate with no detectable social influence between agents, no behavioral convergence over time, and no emergence of turn-taking patterns. Critically, these agents \emph{demonstrated} social behavior capability in structured lab conditions, implying that the environmental difference—not a capability difference—explains the behavioral divergence.

The MAST failure taxonomy \citep{xiao2025mast} identifies coordination failure—the absence of norm-based behavioral coordination—as the dominant failure mode in multi-agent LLM systems (36.9\% of failures). Their finding that structural network topology moderates coordination failure aligns with our environmental design hypothesis and suggests that infrastructure interventions may address the most common failure mode at low cost.

Recent work has also identified risks from competitive incentives overriding prosocial behavior. \citeA{demooy2026} documented that open platform competitive dynamics can amplify tool-use optimization at the expense of social welfare—directly relevant to the task framing (F3) dimension of our experiment.

\subsection{Constitutional Evolution and Emergent Identity}
\label{sec:constitution}

The Constitutional Evolution framework \citep{constitutional2026} provides evidence that evolutionary pressure—environmental selection—can discover optimal behavioral strategies in agent populations independent of individual model capability. This supports our environmental hypothesis and suggests a mechanism (selection pressure) distinct from social learning. Their finding that excessive communication can be counterproductive aligns with our F2 (communication structure) results.

Critically, \citeA{piedrahita2025free} demonstrated that reasoning-capable models may exhibit \emph{degraded} cooperative behavior relative to non-reasoning models, with guilt-aversion mechanisms failing under certain framings. This introduces a potential moderating role for model capability that our study does not fully address—a limitation discussed in Section \ref{sec:limitations}.

\subsection{Theoretical Integration}
\label{sec:theory}

Our Norm Emergence Environmental Model (NEEM) synthesizes these prior findings into a unified framework. NEEM posits that four environmental dimensions—feedback structure (F1), communication topology (F2), task framing (F3), and population stability (F4)—jointly determine whether behavioral convergence crosses the threshold for norm emergence. This framework integrates social translucence theory (F2 infrastructure), social learning theory (F1 feedback), institutional theory (F3 framing), and evolutionary dynamics (F4 selection) into a testable model.

Our primary advance over prior work is the systematic factorial manipulation of all four dimensions simultaneously, enabling detection of both main effects and interactions—something no prior study of LLM norm emergence has achieved.
